---
title: "p8105_hw2_zl3804.Rmd"
author: "Zhuoxuan Li (Eva)"
date: "2026-10-06"
output: github_document
---
```{r}
library(tidyverse)
```
# Problem 1

## Read, clean and describe

```{r}


options(lifecycle_verbosity = "quiet")

NYC_df = read.csv(
  "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  na = c("",".","NA")
  ) |>
  janitor::clean_names()|>
  select(line, station_name, station_latitude, station_longitude, route1:route11, entry, vending, entrance_type, ada)|>
    mutate( entry =
      case_match(
        entry,
        "YES" ~ TRUE,
        "NO" ~ FALSE
      )
    )
```

This NYC subway dataset from MTA contains line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance information.

So far, I have read the original dataset, defined the missing data as NA, used janitor to make variable names cleaner, selected the variables wanted as mentioned, and converted the entry from character to a logical variable[ YES became TRUE, NO became FALSE].

Now I got a dataset called NYC_df with `r nrow(NYC_df)` rows and `r ncol(NYC_df)` columns.

These data are not fully tidy.
Each row represents one station entrance/exit, and most variables form their own columns.
However, the routes served by each station are spread across multiple columns (route1 to route11). According to tidy data principles, each variable should form a single column. Having the same type of information (route) repeated across many columns violates this rule.

## Answering relative questions
```{r}
n_station = 
  NYC_df |> 
  distinct(line, station_name) |> 
  nrow()

n_ada =
  NYC_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

n_novending=
  NYC_df |>
  filter(vending == "NO") |>
  nrow()
```

Q1: There are `r n_station` distinct stations(identified by name and line);

Q2: There are `r n_ada` stations are ADA compliant;

Q3: The proportion of station entrances / exits without vending allow entrance is: `r NYC_df |> filter(vending == "NO") |> summarise(prop = mean(entry, na.rm = TRUE))`.

## Reformat data

```{r}
NYC_tidy_df = 
  NYC_df |>
  mutate(across(route1:route11, as.character)) |>   
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route_name"
  )

n_Atrain = 
  NYC_tidy_df |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()
  
n_Atrain_ada = 
  NYC_tidy_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```

There are `r n_Atrain` serves A train;

Of the stations that serve the A train, `r n_Atrain_ada` are ADA compliant.

# Problem 2

```{r}
library(readxl)

mrwheel_df  =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N743",
    na = c("",".","NA")) |>
  janitor::clean_names() |>
  drop_na(dumpster)

 

mrwheel_df =
  mrwheel_df |>
  mutate(
  sports_balls = round(sports_balls) |>
    as.integer(),
  wheel_name = "mr_trash_wheel",
  year = as.numeric(year)) |>
  relocate(wheel_name) 


```

Combine multiple sheets:

```{r}
professor_df  =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M140",
    na = c("",".","NA")) |>
  janitor::clean_names()

professor_df = drop_na(professor_df, dumpster) |>
  mutate( wheel_name = "professor_trash_wheel",
          plastic_bottles = as.numeric(plastic_bottles)) |>
  relocate(wheel_name) 


gwy_df  =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L395",
    na = c("",".","NA")) |>
  janitor::clean_names()

gwy_df = drop_na(gwy_df, dumpster) |>
  mutate( wheel_name = "gwynnda_the_good_wheel") |>
  relocate(wheel_name) 



joint_wheel = bind_rows( mrwheel_df,professor_df,gwy_df)



```

In the binded dataset, here are `r nrow(joint_wheel)` observations with key variables such as: wheel_names, date, weight_tons, volume_cubic_yards, homes_powered,and quantity of different type of trash, etc.

Of the three wheels, Professor Trash Wheel collected `r sum(pull(professor_df,weight_tons))` tons trash. And the total number of cigarette butts collected by Gwynnda in June of 2022 is `r gwy_df|> filter( year == 2022, month == "June") |> pull(cigarette_butts)|> sum()`

# Problem 3

## Basic demographic information

```{r}
baseline_df = 
  read.csv("data/MCI_baseline.csv",
           na = c("",".","NA"),
           skip = 1) |>
  janitor::clean_names() |>
  filter(is.na(age_at_onset) | age_at_onset > current_age) |>
  mutate(sex=
           case_match(
             sex,
             1 ~ "male",
             0 ~ "female"
           ),
         apoe4=
           case_match(
             apoe4,
             1 ~ "apoe4_carrier",
             0 ~ "apoe4_noncarrier"
           ))
  
```
Discuss important steps in the import process and relevant features of the dataset.

To get clean and tidy baseline dataset, I imported the original data, omitting the first line which was just a instruction note and defining the NAs to drop any participants who do not meet the stated inclusion criteria later. After cleaning the name into snake_case, I removed participants with age_at_onset = age at the study baseline using filter function. Lastly, I used mutate and case_match to ensure that sex and APOE4 carrier status are appropriate encoded.

```{r}
original_baseline = 
  read.csv("data/MCI_baseline.csv",skip = 1) |>
  janitor::clean_names()


```
Here are some key information in this data: Age at the study baseline ,Years of education,gender,Age at the onset of MCI,and wheter carry Apoe4.
We can see here are `r nrow(original_baseline)` participants, among which `r nrow(baseline_df)` develop MCI. 

The average baseline age is `r original_baseline|> pull(current_age)|>mean(na.rm = TRUE)` years old.

`r original_baseline |> filter(sex == 0) |> summarise(prop = mean(apoe4 == 1, na.rm = TRUE))` of women in the study are APOE4 carriers.

## longitudinally observed biomarker values

```{r}
amyloid_df = 
  read.csv("data/mci_amyloid.csv",
           na = c("",".","NA"),
           skip = 1) |>
  janitor::clean_names() 

amy_tidy_df = 
  amyloid_df |>
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    names_prefix = "time_",
    values_to = "biomark_value"
  ) |>
  mutate(
    time = replace(time, time == "baseline", "0"))
  
```
comment on the steps on the import process and the features of the dataset.
Imported mci_amyloid.csv,defined missing data, omitted the first row[note], and turned the variable names into snake_case. Importantly, I need to convert this dataset to a tidy form so I used pivot_longer() to put several times into same variable,making data longer.Meanwhile,in order to make data more pretty, I removed preflix and changed "baseline" into "0".

ORIGINALLY,In INITIAL dataset, we have `r nrow(amyloid_df)` participants with study Id, times elapsed since the study baseline to the visit where biomarker Amyloid _ 42/40 ratio was measured from baseline to 8th follow-up. 

After pivot, we have `r nrow(amy_tidy_df)` observations with study Id, times elapsed since the study baseline to the visit where biomarker Amyloid _ 42/40 ratio was measured from baseline to 8th follow-up and their biomark values.

## Join two dataset

This are observations appearing in only the baseline datasets:

```{r}
onlybaseline_df = anti_join(baseline_df, amy_tidy_df, by = c("id" = "study_id"))
```
Here are `r nrow(onlybaseline_df)` observations in only the baseline datasets. Attention: here I use the tidy baseline dataset which means I excluded those who do not meet the stated inclusion criteria (i.e. no MCI at baseline).

This is observations appearing in only the amyloid datasets:
```{r}
only_amy_df = anti_join(amy_tidy_df, baseline_df, by = c("study_id" = "id"))
```
Here are `r nrow(only_amy_df)` observations in only the amyloid datasets. Attention: here I use the tidy amyloid biomark dataset which means the number of rows/observations does not equal to the number of participants.

`r nrow(onlybaseline_df)` != `r nrow(only_amy_df)` ,so there are participants only in one dataset.

This is participants appearing in both baseline and the amyloid datasets:
```{r}
combine_df = inner_join(baseline_df, amy_tidy_df, by = c("id"="study_id"))
write_csv(combine_df, "data/mci_combined.csv")
```

In the combined_df, here are `r nrow(combine_df)` observations. But need to pay attention to that this combine_df is a combination of two clean and tidy datasets which means in combine dataset, those who do not meet the stated inclusion criteria (i.e. no MCI at baseline) are excluded. Also, the apparence of combine_df is longer form because what got joined is tidy amyloid datasets rather than initial wide one.

Among `r nrow(combine_df)` observations, here are `r length(unique(pull(combine_df,id)))` different unique id.
